A Large-scale Study of Polymorphism in R

### 2019-03-21 paper/main.tex
minor
types

\documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\usepackage{booktabs,listings,xspace}
\lstset{language=R}
\usepackage{my_style}

@@

@@

\begin{figure}{\small
\begin{lstlisting}[style=R]
> normSum <- function( m )  round( m / sum(m), 2)
> normSum(c(1L,3L,6L))
[1] 0.1 0.3 0.6
> normSum(c(1.1,3.3,6.6))
[1] 0.1 0.3 0.6
> normSum(c(1.6,3.3,6.1))
[1] 0.15 0.30 0.55
> normSum(complex(r=rnorm(3),i=rnorm(3)))
[1] 0.49+0.21i 0.30-0.18i 0.22-0.03i
\end{lstlisting}}
\caption{Sample R code}\label{sample}
\end{figure}

@@

\begin{figure}[!ht]
\begin{tabular}{l|l}\hline
\multicolumn{2}{l}{\bf Vectorized data types:}  \\\hline
\k{logical}  & vector of boolean values\\
\k{integer}   & vector of 32 bit integer values\\
\k{double} & vector of 64 bit floating point values\\
\k{complex} & vector of complex values\\
\k{character} & vector of strings values\\
\k{raw} & vector of bytes\\
\k{list} & vector of values of any type\\\hline
\multicolumn{2}{l}{\bf Scalar data types:}\\\hline
\k{NULL}  &  singleton null value\\
\k{closure} & a function with its environment\\
\k{environment} & a mapping from symbol to value \\\hline
\multicolumn{2}{l}{\bf Implementation data types:}\\\hline
&\k{special},
\k{builtin},
\k{symbol},
\k{pairlist},
\k{promise},
\k{language},
\k{char},
\k{...}\\
&\k{any},
\k{expression},
\k{externalprt},
\k{bytecode},
\k{weakref}\\\hline
\end{tabular}\caption{Builtin Types}\label{types}\end{figure}

The first distinction that can be made is based on the value of the
\code{typeof} function.  This function, given a value, will return one of
the following types: \code{logical} (vector of booleans), \code{integer}
(vector of integers), \code{double} (vector of floating points),
\code{complex} (vector of complex numbers), \code{character} (vector of
character strings), \code{raw} (vector of bytes) and \code{list}
(polymorphic vector), \code{NULL} (vector of NULL),
\code{closure} (function), \code{environment} (naming environment),
\code{S4} (S4 object) and others that are used within the implementation of
R (\code{special}, \code{builtin}, \code{symbol}, \code{pairlist},
\code{promise}, \code{language}, \code{char}, \code{...}, \code{any},
\code{expression}, \code{externalptr}, \code{bytecode} and \code{weakref}).

%% TODO : make the types all bold.
\AT{Some examples of use of \code{typeof}:} To build a sense of how these
types arise in R code, consider the following: \AT{We could think of another
  way to do this also.}

\begin{itemize}
\item \code{typeof(5) == typeof(c(5)) == typeof(c(5, 5)) == "double"}.  As
  we mentioned, R is vectorized, so even apparent scalars are vectors.
  \AT{Just for fun,} note that \code{5[1] == 5} in R.
\item to actually get an integer, you need \code{typeof(5L) == "integer"}.
  Doubles are by far the most common numeric type in R \AT{I can get data
    for this}, and you can even use them to index lists and vectors:
  \code{list(1, 2)\[\[1.9\]\] == 1}.  Note the double brackets for list
  indexing.
\item NAs are common in data-centric languages (e.g. MATLAB, \AT{others?}),
  and R is no exception.  As for their type, we see that \code{typeof(NA) ==
    "logical"}, but that isn't to say that NA only has type logical.
  Consider \code{typeof(c(1, NA, 3)\[2\])}: NA actually inhabits every type
  in R, \AT{details}.
\item \AT{other examples?}
\end{itemize}

@@

  - dim

class:

@@

Our corpus is a subset of CRAN, consisting of \PACKAGES packages, accounting
for some \PERCENTCRAN of all packages.  These packages have a total of \RLOC
lines of R code and \CLOC lines of C code. For each package, we extracted
all executable code snippets from documentation, vignettes and tests and ran
them independently recording all call to R functions.  It is noteworthy that
in order to run the scripts in one package, it is often necessary to load a
number of other packages.  In~\cite{issta18}, the authors estimated code
coverage to be around 60\% when including reverse dependencies.  The
rationale for our choice of packages was pragmatic, our infrastructure adds
overhead to the execution of scripts, we eliminated any package that would
take more than one hour to run. We expect to extend our coverage of CRAN for
the next version of this paper.

>>>>

\documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\usepackage{booktabs,listings,xspace,wrapfig}
\lstset{language=R}
\usepackage{my_style}

@@

\newpage  %%Leave here

@@

\begin{figure}[!hb]{\small
\begin{lstlisting}[style=R]
> normSum <- function( m )  round( m / sum(m), 2)
> normSum(c(1L,3L,6L))
[1] 0.1 0.3 0.6
> normSum(c(1.1,3.3,6.6))
[1] 0.1 0.3 0.6
> normSum(c(1.6,3.3,6.1))
[1] 0.15 0.30 0.55
> normSum(complex(r=rnorm(3),i=rnorm(3)))
[1] 0.49+0.21i 0.30-0.18i 0.22-0.03i
\end{lstlisting}}
\caption{Sample R code}\label{sample}
\end{figure}

@@

\begin{wrapfigure}{r}{6.1cm}
\footnotesize\begin{tabular}{l|l@{}}\hline
\multicolumn{2}{l}{\bf Vectorized data types:}  \\\hline
\k{logical}  & vector of boolean values\\
\k{integer}   & vector of 32 bit integer values\\
\k{double} & vector of 64 bit floating points\\
\k{complex} & vector of complex values\\
\k{character} & vector of strings values\\
\k{raw} & vector of bytes\\
\k{list} & vector of values of any type\\\hline
\multicolumn{2}{l}{\bf Scalar data types:}\\\hline
\k{NULL}  &  singleton null value\\
\k{S4}    &  instance of a S4 class \\
\k{closure} & a function with its environment\\
\k{environment} & a mapping from symbol to value \\\hline
\multicolumn{2}{l}{\bf Implementation data types:}\\\hline
\multicolumn{2}{l}{\k{special},
\k{builtin},
\k{symbol},
\k{pairlist},
\k{promise}}\\
\multicolumn{2}{l}{
\k{language},
\k{char},
\k{...}, 
\k{any},
\k{expression},
}\\
\multicolumn{2}{l}{
\k{externalprt},
\k{bytecode},
\k{weakref}}\\\hline
\end{tabular}\caption{Builtin Types}\label{types}\end{wrapfigure}

Figure~\ref{types} lists all of the builtin types that are provided by the
language. They are the possible return values of \k{typeof}. There is no
intrinsic notion of subtyping in R. But, in many context a \k{logical} will
convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
off conversion can occur in corner cases, such as \k{1<"2"} holds and
\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
converted to an integer. R does not distinguish between scalars and vectors
(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
distinguished missing value denoted by \code{NA}. The default type of
\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.

With one exception all vectorized data types are monomorphic, the exception
is the \k{list} type which can hold values of any other type including
\k{list}. For all monomorphic data types, attempting to store a value of a
different type will cause a conversion. Either the value is converted to the
type of vector, or the vector is converted to the type of the value.

Scalar data types include the distinguished \k{NULL} value, which is also of
type \k{NULL}, instance of classes written using the S4 object system,
closures and environments.  The implementation of R has a number of other
types that are mostly not used by user code, they are listed in
Figure~\ref{types} for reference.

@@

@@

Our corpus is a subset of CRAN, consisting of \PACKAGES packages, accounting
for some \PERCENTCRAN of all packages.  These packages have a total of \RLOC
lines of R code and \CLOC lines of C code. For each package, we extracted
all executable code snippets from documentation, vignettes and tests and ran
them independently recording all call to R functions.  It is noteworthy that
in order to run the scripts in one package, it is often necessary to load a
number of other packages.  In~\cite{issta18}, the authors estimated code
coverage to be around 68\% when including reverse dependencies.  The
rationale for our choice of packages was pragmatic, our infrastructure adds
overhead to the execution of scripts, we eliminated any package that would
take more than one hour to run. We expect to extend our coverage of CRAN for
the next version of this paper.
